OpenMPIR

### 2017-10-08 paper.tex
Title changes?
Kei is done

When writing task-parallel programs today, there is a large selection of
potential programming models and implementations to consider. Unfortunately, 
despite having some significant overlaps in semantics, parallel programming
models such as OpenMP~\cite{openmp}, Cilk~\cite{cilk}, Kokkos~\cite{kokkos},
HPX~\cite{hpx}, Charm++~\cite{charm}, Qthreads~\cite{qthreads},
pthreads\cite{pthreads}, MPI~\cite{mpi}, Chapel~\cite{chapel}, UPC~\cite{upc},
OpenCL~\cite{opencl}, and OpenACC~\cite{openacc}, among others, have limited ability to
interoperate either at the level of compile-time analysis or at run-time
execution. In addition, many of these programming models are implemented as
runtime libraries with only primitive compiler support. As a consequence, most are
not able to take advantage of potential compiler analysis and optimization
capabilities.  While there has been some recent work on specializing
compilers to reason about parallel programming libraries~\cite{Moss_2016},
fragmentation among models presents a significant challenge for compiler
writers, which must choose a single model, or multiple intermediate forms, for
analysis and optimization. An ideal solution would be a 
common representation of parallel semantics and constructs.

@@

Compilers implement the majority of their analyses and optimizations on
intermediate representations \textbf{(IRs)} that allow such transformations
to be written once with relative ease and apply to a variety of source
front-end languages such as C or C++.  Historically, the IR of mainstream
compilers such as LLVM~\cite{lattner2004llvm} or GCC's
Gimple~\cite{merrill2003generic} haven't supported parallel constructs. As a
consequence, compilers using these IRs haven't had the ability to reason
about parallelism. The recent work of Schardl et al.\ has shown
that the LLVM IR can be extended to represent fork-join parallelism without
requiring a major rewrite~\cite{tapir}.  They did so by extending the LLVM instruction set
with three instructions capable of representing fork-join parallelism. To
benchmark performance and test accuracy they wrote a front end which translated
programs written with Cilk, a parallel programming model for C/C++, into the
Tapir IR. They also implemented a back end to \textbf{lower} Tapir programs to
vanilla LLVM IR with embedded Cilk runtime calls. They suggested that a similar
approach could be taken with OpenMP and other such frameworks.

@@

The remainder of the paper is organized as follows.  In Section~\ref{Sec:Background} 
we give background and motivation for the presented work. Following that, we describe OpenMP
tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
evaluation setup, including which implementations we compare to and how. We discuss
the results of evaluation in Section~\ref{Sec:Results}, including possible
explanations for discrepancies. We discuss the implications of the results and the
significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
we cover some of the many ways the work could be extended in
Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.

@@

LLVM has gained traction in the broad community for its simplicity,
relative ease of both experimentation and modifications, and extensibility~\cite{lattner2004llvm}.
Historically, LLVM has only had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallel constructs as thin wrappers for calls into run-time
systems. These calls are infeasible for the compiler to reason about because of
the complexity and flexibility of the runtimes being called, and the challenge
of reasoning about the corresponding semantics in a uniform fashion. The
addition of parallel constructs were recently explored by Schardl et
al.~\cite{tapir}. By adding three first-class instructions to LLVM, Tapir
enables analyses and optimizations of parallel constructs. We briefly describe
the key aspects of Tapir in Section~\ref{Sec:Tapir}.
% I removed this as the reference was already given...   PM For a more complete
% description see \cite{tapir}.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
mpi}.  From libraries, to language extensions, to standalone languages, to
combinations of these, the fragmentation of parallel programming models is a
problem that is only getting worse. This means that when writing tooling,
optimizations, or analyses, one much choose which model to target as well as
understand points of contention and interoperability between them. A common
difficulty among all of these is that reasoning about parallel programs is hard
and compilers can potentially help reduce this burden. This is exactly
analogous to the problem LLVM solves: compiler optimizations are hard, so
sharing them is valuable. Tapir is a demonstration that the LLVM approach to
compilation of serial code can be effectively extended to help reason about
parallel code.

@@

Addressing these concerns is a major motivation for our use of Tapir.  By
standardizing an IR, in addition to easing the development of optimizations,
one avoids duplicating work across different implementations. It also enables
the possibility of multiple programming models coexisting peacefully. We will
return to this consideration later in Section~\ref{Sec:Discussion} when
discussing future work.

\section{Tapir} \label{Sec:Tapir}
Tapir is an extension to LLVM proposed by Schardl, Moses, and Leiserson
seeking to resolve issues in optimizing parallel code. Prior to
the introduction of Tapir compilers would represent parallel programs
by directly translating their syntax to opaque runtime calls. This allowed
compilers to support parallel programs but meant that traditional 
optimizations such as code motion or common-subexpression elimination weren't
able to reason about parallel programs. This often led to parallel programs
running significantly slower than expected. In the Tapir project the authors
show that it is possible to represent and optimize fork-join parallel programs
with relatively minimal modifications to the compiler, namely the introduction
of three instructions designed to interface well with an existing compiler.

@@

\caption[Simplified compilation of the unoptimized search code from \figref{search}.]
{Simplified compilation of the unoptimized search code from~\figref{search}.  
\subfigcap{a}~The function \code{search}. The parallel tasks are moved into their own 
functions and accompanying closures. These functions are then passed to OpenMP runtime 
calls. This system obfuscates the program, making it infeasible for an optimizer to 
reason about what is happening.}
  \label{fig:runtime_calls}
\end{figure}

\subsection{Tapir Internals}
Tapir is a compelling solution to this problem because it allows existing
analysis and optimizations to work on parallel programs without significant modifications.
To properly describe Tapir requires a brief introduction to key aspects of LLVM.
LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
sequences of instructions. Blocks always end with terminator instructions that
define the edges of the control flow graph. For example, \figref{CFG} shows a simple
function and its corresponding LLVM code.

@@

Recursive task creations and synchronizations using the constructs result in 
an implicit directed acyclic graph (DAG) that allows both reasoning about and 
visualization of the program execution.  \figref{fib-graph-to-schedule}
shows some example code, a view of the task DAG, and a simplified execution 
schedule mapping the tasks to a team of two threads.  In the example, the Nth 
Fibonacci number is calculated by recursively generating tasks to calculate 
the (N-1)th and (N-2)th Fibonacci numbers.  The \texttt{taskwait} ensures that 
the child tasks have completed before their answers are combined to yield the 
final result.

@@

While we did implement the codegen for the \texttt{task} and
\texttt{taskwait} constructs, it's worth noting that even for these pragmas
the implementation is incomplete. Currently, any clauses modifying the behavior
of the pragmas is ignored. Probably the most common semantics this will change
is for the semantics of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
Surprisingly, this had little effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this problem would
marginally increase performance by reducing the number of memory copies for
variables declared \texttt{private} in OpenMP code.  The fact that behaviour
wasn't changed significantly is worth revisiting, and we do so in
Section~\ref{Sec:Discussion}.

The work represented in this paper is only front-end implementation, which
means we use the only existing back end for Tapir. The existing back end
generates code that calls the Cilk runtime \texttt{libcilkrts}. This has
implications for adhering to the OpenMP specification. For example, environment
variables such as \texttt{OMP\_NUM\_THREADS} are ignored and replaced by
\texttt{CILK\_NWORKERS}.  We discuss some of these issues below, and return to
address others in the discussion of future work.

@@

Upon entering the \texttt{else} branch, the code immediately detaches the basic
block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
a nice example of code that will be optimized by Tapir. The LLVM code
shown in \figref{tapir-example} is not optimized. With optimizations turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

@@

\item \texttt{Alignment}: aligns all protein sequences from  an  input

@@

\item \texttt{FFT}: computes the one-dimensional Fast Fourier Transform

@@

\item \texttt{Fibonacci}: computes the $n$th Fibonacci number using a  recursive
parallelization. While  not  representative  of  an efficient  fibonacci

@@

\item \texttt{Floorplan}: kernel computes the optimal floorplan distribution

@@

\item \texttt{Health}: simulates the Columbian Health Care System. It

@@

\item \texttt{NQueens}: computes  all  solutions  of  the n-queens

@@

\item \texttt{Sort}: sorts a random permutation of $n$ 32-bit numbers with  a

@@

\item \texttt{SparseLU}: computes an LU matrix factorization over

@@

\item \texttt{Strassen}: algorithm  uses  hierarchical  decomposition of a

@@

We set a time-out of 10 minutes, as at least one implementation was always
finishing within 10 seconds, anything running more than 60 times slower becomes
irrelevant.  Because of time constraints correctness checks were not performed on
every run so it is possible some non-determinism was missed.

@@

In this section we discuss the results of running our evaluation on our
implementation, and how its performance compared to the existing
implementations listed in Section~\ref{Sec:Evaluation}. See
Figures~\ref{fig:results} and \ref{fig:results2} for a simple visualization of
the performance. Each graph represents a single benchmark from the previously
enumerated Barcelona benchmarks, and each bar represents one of the
implementations.

@@

As mentioned earlier on \texttt{Floorplan} our implementation returned correct
results nondeterministically because of not having an implementation of the OpenMP
\texttt{critical} and \texttt{atomic} pragmas. While ICC and Clang finished in
roughly 1-2 seconds, the Tapir implementation was finishing in 0.02 seconds
and computing the correct result roughly half of the time. This raises interesting
questions on the cost/benefit relation of the missing OpenMP pragmas.

@@

In this section we investigate \emph{why} performance varies in the ways it
does. While we have yet to identify every discrepancy in performance, we hope
to get some ideas for why the performance of our implementation is generally
better and where each of the implementations is spending the bulk of the
execution time. Our primary tool for this task is profiling. It's worth noting
that lots of information is difficult to infer from profiling, such as sources
of contention, reasons for memory locality, etc.

It is worth noting again that the scope of the work that this paper is
responsible for is only the front end. All performance gains attributable to
the existing Tapir back end and Cilk runtime represent prior work.  We can
break reasons for performance discrepancies into a few categories:
\begin{itemize}
\item Front-end code generation;
\item IR Optimizations; and
\item Runtime efficiency.
\end{itemize}
While we would like to distinguish between these, it is difficult to do so with
only profiling information. A more thorough investigation would require deep
knowledge of each of the front ends and runtimes. We hypothesize that much of
the disparity present in these results is more a function of runtime
implementation differences. As justification for this hypothesis, consider the
\texttt{Fibonacci} benchmark. There are no memory accesses to the heap, and the
code leaves little room for optimization (algorithmic rewrites aside), so it is
effectively a benchmark of how fast a runtime can execute fork-join
parallelism.

@@

There are two cases we examine using the profiler. The first is the case when
there is a large difference in run times between implementations. For these our
hypothesis is that the slower implementations are for some reason spending more
time in the runtime. As mentioned, it is difficult to understand exactly why an
implementation is spending more time in the runtime without further
investigation, but it is a useful sanity check nonetheless. For this case, we
turn to the \texttt{Fib} benchmark.  Indeed, we get roughly the following
breakdowns of work-to-overhead ratio according to the Linux \texttt{perf} tool.
\begin{itemize}
\item \texttt{tapir: ~30\% runtime overhead, ~50\% work, ~20\% other}
\item \texttt{clang: ~85\% runtime overhead, ~5\% work, ~10\% other}
\item \texttt{icc: ~85\% runtime overhead, ~5\% work, ~10\% other}
\item \texttt{gcc: ~100\% runtime overhead, ~0\% work, ~0\% other}
\end{itemize}
This supports our hypothesis that much of the slowdown incurred is overhead in
the runtime. For Tapir the overhead is relatively small given the size of the
work chunks, for \texttt{clang} and \texttt{icc} it is significantly larger,
and for \texttt{gcc} it is overwhelming. It's worth noting that \texttt{clang} 
and \texttt{icc} have effectively the same runtime, so the fact that their 
overheads are similar is unsurprising.

@@

Another way we can discriminate the cause of performance discrepency is by disabling 
Tapir optimizations. This allows us to directly measure the impact that Tapir
optimizations are having on runtime. For \texttt{fibonacci}, for example,
disabling optimizations results in a roughly ~2x performance loss. This is
likely due to Tapir inlining the second task call and moving some memory
operations into registers. This is strong evidence that Tapir optimizations are
an important factor in the performance of compiled task-parallel code. Again,
note that this compounds with runtime performance discrepency to result in the
measured performance reported in Section~\ref{Sec:Results}.

@@

In this section we discuss the relevance of the results and in the context of the
literature. As discussed in Section~\ref{Sec:Introduction}, 
there have been many programming models for implementing parallel programs. While
this paper does not take a stance on which one \emph{should} be used, it does
accept that OpenMP is currently/potentially the most widely used model that has 
a notion of tasks. The goal of this work is to show that a single intermediate
representation can be used to implement multiple programming paradigms with
good performance. Schardl et al.\ showed that Tapir works well as an IR target
for Cilk and surmised that it could be an effective IR for a large subset of
  OpenMP~\cite{tapir}. This work has taken the first step in that direction.

One interesting aspect of this work is that it compiles a language extension
intended for one runtime (OpenMP) to use another (the Cilk runtime). While in
many ways this is a weakness of this work, it does open the possibility of
utilizing this in a more general way to address many significant challenges
such as performance portability. Having a parallel-aware IR enables the
possibility of multiple high-level parallel languages and multiple parallel
runtime targets in the same way that LLVM does for serial code. While the
OpenMP specification's requirement to have access to runtime routines
complicates things, for many programming models it seems at least conceivable
to map them onto multiple different runtimes. This opens the possibility of
mixing programming models at compile time by targeting the same runtime. For
example, if both OpenMP and Cilk targeted Tapir instructions, one could have a
large application where some parts were written using Cilk, while others used
OpenMP, and the result would not suffer the incompatibility problems often
seen today.

The surprising fact that program behavior wasn't changed significantly is worth
discussing. We surmise that it is likely due to the semantics of variables
borrowed from the Cilk codegen and their relation to the OpenMP specification
of semantics. In particular, \emph{OpenMP specifies a relaxed memory model
for shared variables}, in which writes by one thread need only be visible at
to other threads at explicit or implicit flushes. 
This results in the Cilk implementation of only writing
to a shared variable at the end of a parallel section being a valid
implementation of OpenMP's specification. In contrast, existing OpenMP
implementations sometimes write shared variables to memory on every write
access. This is a potential for performance discrepancy because of differing
implementation behavior.
%Stephen:This topic is not discuessed later, but was touched on before --
%moved?  and we will return to it when discussing performance on the Barcelona
%OpenMP task suite. 
There is of course the fact that technically, by copying the value of variables
back to the surrounding context even for \texttt{private} variables, one isn't
following the specification. We surmise that in practice this hasn't had an
effect because \texttt{private} variables are often used for inputs to
calculations that are not accessed after the tasks complete, or for performance
rather than their semantic properties.  Also, \texttt{firstprivate} is the
default data-sharing attribute for tasks.

@@

For two examples, consider the problematic \texttt{atomic} and
\texttt{critical} pragmas from the \texttt{Floorplan} benchmark. The
\texttt{atomic} pragma would likely not be too difficult to translate from
existing OpenMP codegen. One would need to simply maintain and then modify the
contents of the original stack pointer to the shared variable. The
\texttt{critical} section would likely be a little more difficult. Still, it
should be possible to modify existing OpenMP code to block on that section.

Still, whether or not it's \emph{possible} to implement these extra features
is a separate question from whether or not it's \emph{useful}. If implementing
extra features require using special synchronization primitives, or calls into
runtimes, or any other tool other than the Tapir instructions, then the
compiler cannot reason about code using those features, and many of the
advantages we've discussed become moot, e.g., backend compatibility, generic 
optimization, etc. Features like the OpenMP \texttt{atomic} pragma don't 
require any changes to control flow, and therefore are likely to work well 
with Tapir analyses and optimizations. On the other hand features like OpenMP's
\texttt{critical} section \emph{do} require changes to control flow that seem
challenging to map onto Tapir's instructions. Furthermore, even if it
is possible to implement features using the Tapir instructions, it may require 
such radical entangling of the code that even if analysis is technically
possible it becomes infeasible.

@@

While there is general agreement that there needs to be IR support for
parallelism in parallel programs, there isn't consensus on which of these
approaches is best. Approaches that are easier at first, such as metadata
or intrinsic approaches, are similarly easy to extend but can become unwieldy. There is
a reason that it's difficult to add instructions to LLVM: any added IR
construct needs to be considered in many locations. From debugging by printing
out IR, to case analysis for different control flow constructs, having 
a proper set of instructions to denote parallelism has numerous advantages. 
Additionally, any hope of having a formal semantics for an IR depends on a 
simple, concise set of parallelism constructs, something that metadata and
intrinsics approaches can easily avoid.

>>>>

When writing task-parallel programs today, there is a large selection of
potential programming models and implementations to consider. Unfortunately, 
despite having some significant overlaps in semantics, parallel programming
models such as OpenMP~\cite{openmp}, Cilk~\cite{cilk}, Kokkos~\cite{kokkos},
HPX~\cite{hpx}, Charm++~\cite{charm}, Qthreads~\cite{qthreads},
pthreads\cite{pthreads}, MPI~\cite{mpi}, Chapel~\cite{chapel}, UPC~\cite{upc},
OpenCL~\cite{opencl}, and OpenACC~\cite{openacc}, among others, have limited ability to
interoperate either at the level of compile-time analysis or at run-time
execution. Moreover, many of these programming models are implemented as
runtime libraries with only primitive compiler support. As a consequence, most are
not able to take advantage of potential compiler analysis and optimization
capabilities.  While there has been some recent work on specializing
compilers to reason about parallel programming libraries~\cite{Moss_2016},
fragmentation among models presents a significant challenge for compiler
writers who must choose a single model, or multiple intermediate forms, for
analysis and optimization. An ideal solution would be a 
common representation of parallel semantics and constructs.

@@

Compilers implement the majority of their analyses and optimizations on
intermediate representations \textbf{(IRs)} that allow such transformations
to be written once with relative ease and applied to a variety of source
front-end languages such as C or C++.  Historically, the IR of mainstream
compilers such as LLVM~\cite{lattner2004llvm} or GCC's
Gimple~\cite{merrill2003generic} haven't supported parallel constructs. As a
consequence, compilers using these IRs haven't had the ability to reason
about parallelism. The recent work of Schardl et al.\ has shown
that the LLVM IR can be extended to represent fork-join parallelism without
requiring a major rewrite~\cite{tapir}.  They did so by extending the LLVM instruction set
with three instructions capable of representing fork-join parallelism. To
benchmark performance and test accuracy they wrote a front end which translated
programs written with Cilk, a parallel programming model for C/C++, into their
Tapir IR. They also implemented a back end to \textbf{lower} Tapir programs to
vanilla LLVM IR with embedded Cilk runtime calls. They suggested that a similar
approach could be taken with OpenMP and other such frameworks.

@@

The remainder of the paper is organized as follows.  In
Section~\ref{Sec:Background} we give background and motivation for the
presented work. Following that, we describe Tapir in Section~\ref{Sec:Tapir},
and OpenMP tasking in Section~\ref{Sec:OpenMP}.  We then discuss the
implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe
the evaluation setup, including which implementations we compare to and
how. We discuss the results of evaluation in Section~\ref{Sec:Results},
including possible explanations for discrepancies. We discuss the implications
of the results and the significance of this work in a large context
in Section~\ref{Sec:Discussion}. Finally, we cover some of the many ways the work
could be extended in Section~\ref{Sec:Future}, and conclude in
Section~\ref{Sec:Conclusion}.

@@

LLVM has gained traction in the broad community for its simplicity,
relative ease of both experimentation and modifications, and extensibility~\cite{lattner2004llvm}.
Historically, LLVM has only had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallel constructs as thin wrappers for calls into run-time
systems. These calls are infeasible for the compiler to reason about because of
the complexity and flexibility of the runtimes being called, and the challenge
of reasoning about the corresponding semantics in a uniform fashion. The
addition of parallel constructs was recently explored by Schardl et
al.
%~\cite{tapir}.
By adding three first-class instructions to LLVM, Tapir
enables analyses and optimizations of parallel constructs. We briefly describe
the key aspects of Tapir in Section~\ref{Sec:Tapir}.
% I removed this as the reference was already given...   PM For a more complete
% description see \cite{tapir}.
% KD removed it again.  Also, citatations are meta-syntactic.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
mpi}.  From libraries, to language extensions, to standalone languages, to
combinations of these, the fragmentation of parallel programming models is a
problem that is only getting worse. This means that when writing tooling,
optimizations, or analyses, one must choose which models to target as well as
understand points of contention and interoperability between them. A common
difficulty among all of these is that reasoning about parallel programs is hard
and compilers can potentially help reduce this burden. This is exactly
analogous to the problem LLVM solves: compiler optimizations are hard, so
sharing them is valuable. Tapir is a demonstration that the LLVM approach to
compilation of serial code can be effectively extended to help reason about
parallel code.

@@

Addressing these concerns is a major motivation for our use of Tapir.  By
standardizing an IR, in addition to easing the development of optimizations,
one avoids duplicating work across different implementations. It also enables
the possibility of multiple programming models coexisting peacefully. We will
return to this consideration later in Section~\ref{Sec:Future} when
discussing future work.

\section{Tapir} \label{Sec:Tapir}
Tapir is an extension to LLVM
%proposed by Schardl, Moses, and Leiserson
seeking to resolve issues in optimizing parallel code. Prior to
the introduction of Tapir, compilers would represent parallel programs
by directly translating their syntax to opaque runtime calls. This allowed
compilers to support parallel programs but meant that traditional 
optimizations such as code motion or common-subexpression elimination weren't
able to reason about parallel programs. This often led to parallel programs
running significantly slower than expected. In the Tapir project the authors
show that it is possible to represent and optimize fork-join parallel programs
with relatively minimal modifications to the compiler, namely the introduction
of three instructions designed to interface well with an existing compiler.

@@

\caption[Simplified compilation of the unoptimized search code from \figref{search}.]
{Simplified compilation of the unoptimized search code from~\figref{search}.\subfigcap{a}
          %~The function \code{search}.
The parallel tasks are moved into their own 
functions and accompanying closures. These functions are then passed to OpenMP runtime 
calls. This obfuscates the program, making it infeasible for an optimizer to 
reason about what is happening.}
  \label{fig:runtime_calls}
\end{figure}

\subsection{Tapir Internals}
Tapir is a compelling solution to this problem because it allows existing
analysis and optimizations to work on parallel programs without significant modifications.
To properly describe Tapir requires a brief introduction to key aspects of LLVM\@.
LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
sequences of instructions. Blocks always end with terminator instructions that
define the edges of the control flow graph. For example, \figref{CFG} shows a simple
function and its corresponding LLVM code.

@@

Recursive task creation and synchronization using the constructs result in 
an implicit directed acyclic graph (DAG) that allows both reasoning about and 
visualization of the program execution.  \figref{fib-graph-to-schedule}
shows some example code, a view of the task DAG, and a simplified execution 
schedule mapping the tasks to a team of two threads.  In the example, the $n$th 
Fibonacci number is calculated by recursively generating tasks to calculate 
the ($n-1$)th and ($n-2$)th Fibonacci numbers.  The \texttt{taskwait} ensures that 
the child tasks have completed before their answers are combined to yield the 
final result.

@@

While we did implement the codegen for the \texttt{task} and
\texttt{taskwait} constructs, it's worth noting that even for these pragmas
the implementation is incomplete. Currently, any clauses modifying the behavior
of the pragmas is ignored. Probably the most common semantics this will change
is for the semantics of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
Surprisingly, this had little effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this problem would
marginally increase performance by reducing the number of memory copies for
variables declared \texttt{private} in OpenMP code.  The fact that behavior
wasn't changed significantly is worth revisiting, and we do so in
Section~\ref{Sec:Discussion}.

The work represented in this paper is only front-end implementation, which
means we use the only existing back end for Tapir. The existing back end
generates code that calls the Cilk runtime \texttt{libcilkrts}. This has
implications for adhering to the OpenMP specification. For example, environment
variables such as \texttt{OMP\_NUM\_THREADS} are ignored and replaced by
\texttt{CILK\_NWORKERS}.  We discuss some of these issues following and return to
address others in the discussion of future work.

@@

Upon entering the \texttt{else} branch, the code immediately detaches the basic
block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls \texttt{sync}. This is
a nice example of code that will be optimized by Tapir. The LLVM code
shown in \figref{tapir-example} is not optimized. With optimizations turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

@@

\item \texttt{Alignment} aligns all protein sequences from  an  input

@@

\item \texttt{FFT} computes the one-dimensional Fast Fourier Transform

@@

\item \texttt{Fibonacci} computes the $n$th Fibonacci number using a  recursive
parallelization. While  not  representative  of  an efficient  Fibonacci

@@

\item \texttt{Floorplan} kernel computes the optimal floorplan distribution

@@

\item \texttt{Health} simulates the Columbian Health Care System. It

@@

\item \texttt{NQueens} computes  all  solutions  of  the n-queens

@@

\item \texttt{Sort} sorts a random permutation of $n$ 32-bit numbers with  a

@@

\item \texttt{SparseLU} computes an LU matrix factorization over

@@

\item \texttt{Strassen} uses  hierarchical  decomposition of a

@@

We set a time-out of 10 minutes, as at least one implementation was always
finishing within 10 seconds so anything running more than 60 times slower becomes
irrelevant.  Because of time constraints correctness checks were not performed on
every run so it is possible that some non-determinism was missed.

@@

In this section we discuss the results of running our evaluation on our
implementation, and how its performance compared to the existing
implementations listed in Section~\ref{Sec:Evaluation}. 
Figures~\ref{fig:results} and \ref{fig:results2} give simple visualizations of
performance. Each graph represents a single benchmark from the previously
enumerated Barcelona benchmarks, and each bar represents one of the
implementations.

@@

As mentioned earlier regarding \texttt{Floorplan} our implementation returned correct
results nondeterministically because of not having an implementation of the OpenMP
\texttt{critical} and \texttt{atomic} pragmas. While ICC and Clang finished in
roughly 1-2 seconds, the Tapir implementation was finished in 0.02 seconds
and computed the correct result roughly half of the time. This raises interesting
questions on the cost/benefit relation of the missing OpenMP pragmas.

@@

In this section we investigate \emph{why} performance varies in the ways it
does. While we have yet to identify every discrepancy in performance, we hope
to get some ideas for why the performance of our implementation is generally
better and where each of the implementations is spending the bulk of the
execution time. Our primary tool for this task is profiling, acknowledging
that many kinds information are difficult to infer from profiling, such as sources
of contention, reasons for memory locality, etc.

It is worth noting again that the scope of the work that this paper reports
is only the front end. All performance gains attributable to
the existing Tapir back end and Cilk runtime represent prior work.  We can
separate reasons for performance discrepancies into a few categories:
\begin{itemize}
\item Front-end code generation;
\item IR Optimizations; and
\item Runtime efficiency.
\end{itemize}
While we would like to distinguish between these, it is difficult to do so with
only profiling information. A more thorough investigation would require deep
knowledge of each of the front ends and runtimes. We hypothesize that much of
the disparity present in these results is more a function of runtime
implementation differences. As justification for this hypothesis, consider the
\texttt{Fibonacci} benchmark. There are no memory accesses to the heap, and the
code leaves little room for optimization (algorithmic rewrites aside), so it is
effectively a benchmark of how fast a runtime can execute fork-join
parallelism.

@@

There are two cases we examine using the profiler. The first is the case when
there is a large difference in run times between implementations. For these our
hypothesis is that the slower implementations are for some reason spending more
time in the runtime. As mentioned, it is difficult to understand exactly why an
implementation is spending more time in the runtime without further
investigation, but it is a useful sanity check nonetheless. For this case, we
turn to the \texttt{Fibonacci} benchmark.  We get roughly the following
breakdowns of work-to-overhead ratio according to the Linux \texttt{perf} tool.
\begin{itemize}
\item \texttt{tapir}: ~30\% runtime overhead, ~50\% work, ~20\% other
\item \texttt{clang}: ~85\% runtime overhead, ~5\% work, ~10\% other
\item \texttt{icc}: ~85\% runtime overhead, ~5\% work, ~10\% other
\item \texttt{gcc}: ~100\% runtime overhead, ~0\% work, ~0\% other
\end{itemize}
This supports our hypothesis that much of the slowdown incurred is overhead in
the runtime. For Tapir the overhead is relatively small given the size of the
work chunks, for \texttt{clang} and \texttt{icc} it is significantly larger,
and for \texttt{gcc} it is overwhelming. \texttt{clang} 
and \texttt{icc} have effectively the same runtime, so the fact that their 
overheads are similar is unsurprising.

@@

Another way we can discriminate the cause of performance discrepency is by disabling 
Tapir optimizations. This allows us to directly measure the effect that Tapir
optimizations are having on runtime. For \texttt{fibonacci}, for example,
disabling optimizations results in a roughly ~2x performance loss. This is
likely due to Tapir inlining the second task call and moving some memory
operations into registers. This is strong evidence that Tapir optimizations are
an important factor in the performance of compiled task-parallel code. Again,
note that this compounds with runtime performance differences to give the
measured performance reported in Section~\ref{Sec:Results}.

@@

In this section we discuss the relevance of the results and in the context of the
literature. As discussed in Section~\ref{Sec:Introduction}, 
there have been many programming models for implementing parallel programs. While
this paper does not take a stance on which one \emph{should} be used, it does
accept that OpenMP is currently/potentially the most widely used model that has 
a notion of tasks. The goal of this work is to show that a single intermediate
representation can be used to implement multiple programming paradigms with
good performance. Schardl et al.\ showed that Tapir works well as an IR target
for Cilk and surmised that it could be an effective IR for a large subset of
  OpenMP. This work has taken the first step in that direction.

One interesting aspect of this work is that it compiles a language extension
intended for one runtime (OpenMP) to use another (Cilk). While in
many ways this is a weakness of this work, it does open the possibility of
utilizing this in a more general way to address many significant challenges
such as performance portability. Having a parallel-aware IR enables the
possibility of multiple high-level parallel languages and multiple parallel
runtime targets in the same way that LLVM does for serial code. While the
OpenMP specification's requirement to have access to runtime routines
complicates things, for many programming models it seems at least conceivable
to map them onto multiple different runtimes. This opens the possibility of
mixing programming models at compile time by targeting the same runtime. For
example, if both OpenMP and Cilk targeted Tapir instructions, one could have a
large application where some parts were written using Cilk, while others used
OpenMP, and the result would not suffer the incompatibility problems often
seen today.

The surprising fact that program behavior wasn't changed significantly is worth
discussing. We surmise that it is likely due to the semantics of variables
borrowed from the Cilk codegen and their relation to the OpenMP specification
of semantics. In particular, \emph{OpenMP specifies a relaxed memory model
for shared variables} in which writes by one thread need only be visible at
to other threads at explicit or implicit flushes. 
This results in the Cilk implementation of only writing
to a shared variable at the end of a parallel section being a valid
implementation of OpenMP's specification. In contrast, existing OpenMP
implementations sometimes write shared variables to memory on every write
access. This is a potential for performance discrepancy because of differing
implementation behavior.
%Stephen:This topic is not discuessed later, but was touched on before --
%moved?  and we will return to it when discussing performance on the Barcelona
%OpenMP task suite. 
There is of course the fact that technically, by copying the value of variables
back to the surrounding context even for \texttt{private} variables, one isn't
following the specification. We surmise that in practice this hasn't had an
effect because \texttt{private} variables are often used for inputs to
calculations that are not accessed after the tasks complete, or for performance
rather than their semantic properties.  Also, \texttt{firstprivate} is the
default data-sharing attribute for tasks.

@@

For two examples, consider the problematic \texttt{atomic} and
\texttt{critical} pragmas from the \texttt{Floorplan} benchmark. The
\texttt{atomic} pragma would likely not be too difficult to translate from
existing OpenMP codegen: one would need to simply maintain and then modify the
contents of the original stack pointer to the shared variable. The
\texttt{critical} section would likely be a little more difficult. Still, it
should be possible to modify existing OpenMP code to block on that section.

Still, whether or not it's \emph{possible} to implement these extra features
is a separate question from whether or not it's \emph{useful}. If implementing
extra features require using special synchronization primitives, or calls into
runtimes, or any other tool other than the Tapir instructions, then the
compiler cannot reason about code using those features, and many of the
advantages we've discussed, e.g., backend compatibility, generic 
optimization, etc., become moot, Features like the OpenMP \texttt{atomic} pragma don't 
require any changes to control flow, and therefore are likely to work well 
with Tapir analyses and optimizations. On the other hand, features like OpenMP's
\texttt{critical} section \emph{do} require changes to control flow that seem
challenging to map onto Tapir's instructions. Furthermore, even if it
is possible to implement features using the Tapir instructions, it may require 
such radical entangling of the code that even if analysis is technically
possible it becomes infeasible.

@@

While there is general agreement that there needs to be IR support for
parallelism in parallel programs, there isn't consensus on which of these
approaches is best. Approaches that are easier at first, such as metadata
or intrinsic approaches, are similarly easy to extend but can become unwieldy. There is
a reason that it's difficult to add instructions to LLVM: any added IR
construct needs to be considered in many locations. From debugging by printing
out IR, to case analysis for different control flow constructs, having 
a proper set of instructions to denote parallelism has numerous advantages. 
Additionally, any hope of having a formal semantics for an IR depends on a 
simple, concise set of parallelism constructs, something that metadata and
intrinsics approaches can promptly defeat.
